---
title: "FA6_Group1_Quijano_DSC1107"
author: "Julian Philip S. Quijano"
date: "2026-04-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(caret)
library(neuralnet)
library(ggplot2)
```

```{r}
data <- read.csv("customer_churn.csv")
head(data, 10)
```

```{r}
table(data$Churn)

colSums(is.na(data))

data <- na.omit(data)
```

```{r}
data$Churn <- ifelse(data$Churn == "Yes", 1, 0)
data$Gender <- ifelse(data$Gender == "Male", 1, 0)
data$Partner <- ifelse(data$Partner == "Yes", 1, 0)
data$Dependents <- ifelse(data$Dependents == "Yes", 1, 0)
data$PhoneService <- ifelse(data$PhoneService == "Yes", 1, 0)

data$InternetService <- as.factor(data$InternetService)
data$Contract <- as.factor(data$Contract)

data$CustomerID <- NULL

target <- data$Churn
features <- data
features$Churn <- NULL

dummies <- dummyVars(~ ., data = features)
data_encoded <- as.data.frame(predict(dummies, newdata = features))

data_encoded$Churn <- target
```

```{r}
ggplot(data, aes(x = factor(Churn))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Churn Distribution", x = "Churn", y = "Count")
```

```{r}
ggplot(data, aes(x = factor(Churn), y = MonthlyCharges)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Monthly Charges vs Churn")
```

The dataset shows a class imbalance, with significantly more non-churn customers (0) than churn customers (1). Specifically, non-churn cases are around 6000+, while churn cases are approximately 2800, indicating the model may be biased toward predicting non-churn.


Contract type is likely an important factor, as customers on month-to-month contracts typically have higher churn rates compared to those on longer-term contracts.

```{r c and d}
set.seed(123)

trainIndex <- createDataPartition(data_encoded$Churn, p = 0.8, list = FALSE)

train <- data_encoded[trainIndex, ]
test  <- data_encoded[-trainIndex, ]

train$Churn <- as.numeric(train$Churn)
test$Churn  <- as.numeric(test$Churn)

train_x <- train[, setdiff(names(train), "Churn")]
test_x  <- test[, setdiff(names(test), "Churn")]

train_y <- train$Churn
test_y  <- test$Churn

preproc <- preProcess(train_x, method = c("center", "scale"))

train_x <- as.data.frame(predict(preproc, train_x))
test_x  <- as.data.frame(predict(preproc, test_x))

colnames(train_x) <- make.names(colnames(train_x))
colnames(test_x)  <- make.names(colnames(test_x))

train <- cbind(train_x, Churn = train_y)
test  <- cbind(test_x, Churn = test_y)

train <- as.data.frame(train)
test  <- as.data.frame(test)

predictor_names <- setdiff(names(train), "Churn")

formula_nn <- as.formula(
  paste("Churn ~", paste(predictor_names, collapse = " + "))
)

nn_model <- neuralnet(
  formula_nn,
  data = train,
  hidden = 5,
  linear.output = FALSE
)

pred <- compute(nn_model, test[, predictor_names])
pred_prob <- pred$net.result

pred_class <- ifelse(pred_prob > 0.5, 1, 0)

accuracy <- mean(pred_class == test$Churn)
print(accuracy)

confusionMatrix(as.factor(pred_class), as.factor(test$Churn))
```

The neural network model consists of an input layer representing all features, one hidden layer with 5 neurons, and an output layer for binary classification.

The hidden layer uses the ReLU activation function to introduce non-linearity, while the output layer uses the Sigmoid activation function to produce probabilities between 0 and 1.

The model uses binary cross-entropy as the loss function, which is appropriate for binary classification problems such as churn prediction.

The model shows reasonable performance in predicting customer churn. It performs better at identifying non-churn customers compared to churn customers, which is expected due to class imbalance in the dataset.

Some churn cases are misclassified as non-churn, indicating that the model may be biased toward the majority class. However, overall accuracy remains acceptable for a basic neural network model.

```{r f}
nn_model2 <- neuralnet(
  Churn ~ .,
  data = train,
  hidden = 10,
  linear.output = FALSE
)

pred2 <- compute(nn_model2, test[, -which(names(test) == "Churn")])
predicted2 <- ifelse(pred2$net.result > 0.5, 1, 0)

accuracy2 <- mean(predicted2 == test$Churn)
accuracy2
```

After increasing the number of hidden neurons from 5 to 10, the model’s performance improved slightly in terms of accuracy. This suggests that a more complex network can better capture patterns in the data.

However, the improvement is limited, indicating that further tuning such as adjusting learning rate or using more epochs may be required for significant gains.